Concurrent Programming in Harmony

### 2021-12-15 book/paper.tex
Latest

\begin{code}
\textbf{assert} \texttt{queue.get}(\textit{q}) in \{ \textbf{None}, 1, 2 \}
\end{code}

@@

\begin{figure}
\begin{code}
\harmonysource{queueseq}
\end{code}
\caption{\harmonylink{code/queueseq.hny} Sequential but not a concurrent
queue implementation}
\label{fig:queuebug}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{figures/queuebug1.pdf}
\end{center}
\caption{\harmonyref{queuebug.html}{Harmony output of running \autoref{fig:qtest3} against
\autoref{fig:queuebroken}}}
\label{fig:queuebug1}
\end{figure}

@@

Let us run the code through some of the test programs in the last
chapter.  Harmony does not detect any issues with the sequential test
in \autoref{fig:qtestseq}.
(Run this using the \texttt{-m} flag like this:
\texttt{harmony -m queue=queuebroken code/qtestseq.hny})
The concurrent \texttt{put} test of
\autoref{fig:qtest1} as well as the concurrent \texttt{get} test of
\autoref{fig:qtest2} also find no problems.  However, when we run
the new queue code through the test in \autoref{fig:qtest3}, Harmony
finds a problem.

\autoref{fig:queuebug1} shows the Harmony output of running
the test in \autoref{fig:qtest3} against the queue code
in \autoref{fig:queuebroken}.
There is quite a bit of information in the Harmony output, and it's
important to learn to navigate through it.
Let's start with looking at the red text.  Harmony found a safety
violation (something bad happened during one of the possible executions),
and in particular \texttt{putter}(1) (thread T2) was trying to dereference
\texttt{alloc}.\textit{pool}[0].\texttt{lock} in turn 5.

The \texttt{alloc} module maintains a shared array \textit{pool} that
it uses for dynamic allocation.  Apparently \texttt{putter}(1) tried to access
\textit{pool}[0], but it does not exist, meaning that either it was not
yet allocated, or it had been freed since it was allocated.
When we look at the top half of the figure, we see that in fact
\texttt{putter}(0) (T1) allocated \textit{pool}[0] in turn 2, while
\texttt{getter}() (T4) released it in turn 4.

So how did we get there?  We can start by single stepping through the
actions of \texttt{putter}(0) by clicking on its first block.
By hitting return repeatedly, we can go through the lines of code that
it is executing.  Doing so, we can see that it allocates
\textit{pool}[0] and uses that as the node to contain value 0 and add
this node to the queue \textit{qtest}.  At the end of its turn,
\texttt{putter}(0) has just finished \texttt{put}(0).  Looking at the
state in the top right, everything looks good.  The first node
(\textit{testq}) points to the allocated node with the value 0 in it,
and its \texttt{next} pointer is \textbf{None}.  The locks on both nodes
are released.  So far so good.

>>>>

\begin{code}
\textbf{assert} \texttt{queue.get}($q$) in \{ \textbf{None}, 1, 2 \}
\end{code}

@@

\begin{figure}
\begin{code}
\harmonysource{queueseq}
\end{code}
\caption{\harmonylink{code/queueseq.hny} Sequential but not a concurrent
queue implementation}
\label{fig:queueseq}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{figures/queuebroken1.png}
\end{center}
\caption{\harmonyref{queuebug.html}{Harmony output of running
\autoref{fig:qtestpar} against \autoref{fig:queuebroken}}}
\label{fig:queuebroken1}
\end{figure}

@@

Let us run the code through the test programs in the last
chapter.  Harmony does not detect any issues with the sequential test
in \autoref{fig:qtestseq}.
(Run this using the \texttt{-m} flag like this:
\texttt{harmony -m queue=queuebroken code/qtestseq.hny})
However, when we run
the new queue code through the test in \autoref{fig:qtestpar}, Harmony
finds a problem (even without specifying a behavior).  The command line
for this is \texttt{harmony -m queue=queuebroken code/qtestpar.hny})

Before we go look at the details of what went wrong, we want to make
sure we generate the simplest scenario.  So first we want to explore
what the smallest \texttt{NOPS} (number of operations or number of
threads) that causes the bug to occur.  With some experimentation,
we find that \texttt{NOPS} = 2 does not find a problem, but
\texttt{NOPS} = 3 does
(\texttt{harmony -m queue=queuebroken -c NOPS=3 code/qtestpar.hny})).
\autoref{fig:queuebroken1} shows the Harmony output.

There is quite a bit of information in the Harmony output, and while
it may seem intimidating, we have to learn to navigate through it
step-by-step.
Let's start with looking at the red text.  Harmony found a safety
violation (something bad happened during one of the possible executions),
and in particular \texttt{thread}(2) (thread T2) was trying to dereference
\texttt{?alloc["pool"][0]["lock"]}
(aka \texttt{?alloc}.\textit{pool}[0].\texttt{lock}).

The \texttt{alloc} module maintains a shared array \textit{pool} that
it uses for dynamic allocation.  Apparently T2 tried to access
\textit{pool}[0], but it does not exist, meaning that either it was not
yet allocated, or it had been freed since it was allocated.
When we look at the top half of the figure, we see that in fact thread
T1 allocated \textit{pool}[0] in turn 2, but T3 freed it in turn 4.
Looking back down, we see that T1 executed \texttt{thread}(1) and has
since terminated, while T3 is executing \texttt{thread}(3).

Looking further at the stack traces, we can see that
T3 was in the process of executing
\texttt{release}(?$q$.\texttt{lock}) within
\texttt{get}(?$q$).  Looking in the top green row,
we see that the value of $q$ is
\{ \texttt{.lock}: \texttt{True}, \texttt{.next}: \texttt{None}, \texttt{.value}: \texttt{None} \}.
T1 is currently executing
\texttt{acquire}(\texttt{?alloc}.\textit{pool}[0].\texttt{lock})
within \texttt{put}(?$q$, 2), but 
\texttt{alloc}.\textit{pool}[0] does not exist.
The corresponding line of Harmony code is
\textbf{atomically when not} !\textit{binsema}
in line 25 of the \texttt{sync} module.

So how did we get there?  In the top we can see that the order of
events was the following:
\begin{enumerate}
\item[0:] initialization completed, leaving the value of $q$ at
\{ \texttt{.lock}: \texttt{False}, \texttt{.next}: \texttt{None}, \texttt{.value}: \texttt{None} \};
\item[1:] thread T1 (\texttt{thread}(1)) ran and finished executing
\texttt{put}(1) (see the output column for that clue: the thread printed that).
We can see that $q$.\texttt{next} now points to \texttt{alloc}.\textit{pool}[0],
which the thread must have allocated.  The contents is 
\{ \texttt{.lock}: \texttt{False}, \texttt{.next}: \texttt{None}, \texttt{.value}: 1 \}, as expected;
\item[2:] thread T2 (\texttt{thread}(1)) started running, calling
\texttt{put}(?$q$, 2).  We can see it got as far as putting 2 on the queue;
\item[3:] thread T3 (\texttt{thread}(1)) started running, calling
\texttt{get}($?q$).  We can also see that it freed \textit{pool}[0].
We know that it is in the process of releasing $q$.\texttt{lock};
\item[4:] thread T2 resumes and tries to access \textit{pool}[0], which
no longer exists because T3 just freed it.
\end{enumerate}

Clearly there was a race in which T2 was trying to lock
?\textit{pool}[0].\texttt{lock} (which contained the node with the value 1)
while T3 was freeing that very same node, and T2 lost the race.

We can start by single stepping through the
actions of \texttt{putter}(0) by clicking on its first block.
By hitting return repeatedly, we can go through the lines of code that
it is executing.  Doing so, we can see that it allocates
\textit{pool}[0] and uses that as the node to contain value 0 and add
this node to the queue \textit{qtest}.  At the end of its turn,
\texttt{putter}(0) has just finished \texttt{put}(0).  Looking at the
state in the top right, everything looks good.  The first node
(\textit{testq}) points to the allocated node with the value 0 in it,
and its \texttt{next} pointer is \textbf{None}.  The locks on both nodes
are released.  So far so good.
